Citation Analysis of Computer Science Publications

### 2019-12-10 Scientometrics/comp.tex
update

\institute{Sitaram Devarakonda,  \at
              Netelabs, NET ESolutions Corporation, McLean, VA \\
              %Tel.: +123-45-678910\\
              %Fax: +123-45-678910\\
              \email{sitaram@nete.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
              Dmitriy Korobskiy \at
              Netelabs, NET ESolutions Corporation, McLean, VA  \\
              \email{dk@nete.com}
           \and
           Tandy Warnow \at
              Dept of Computer Science, University of Illinois Urbana-Champaign, Champaign IL \\
              \email{warnow@illinois.edu}
           \and
           George Chacko \at
              Netelabs, NET ESolutions Corporation, McLean, VA \\
              \email{netelabs@nete.com}
}

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification}. All three rely on applying one or more journal-derived labels to articles.  Of these, Scopus and the Web of Science also include global bibliographies with citation links and proprietary unique identifiers for publications, some of which can be matched to digital object identifiers (DOIs). Applying labels designed to represent journals, to articles, can be problematic in terms of accuracy given broad scope in some journals,  and also because fields evolve faster than their journal classifications. [should we put in a comment about why article level labels provided by authors have limited use and that enforcement of controlled vocabularies varies by publisher?] Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017} \textcolor{blue}{need comprehensive list of citations}. Article classification systems have been constructed, however, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need a nother bridging sentence to the next paragraph].

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the XML source file and loaded into a PostgreSQL database. We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings. References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (see Fig~\ref{fig:ar_cp_annotation}).

@@

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, \textcolor{blue}{Sitaram please confirm}, we restricted further analysis to highly cited articles, those in the 90th percentile of citations or higher, resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible ${n \choose 2}$ reference pairs were generated, where $n$ is the number of references in a publication. The cited reference pairs were restricted to those in the set of 212,311 highly cited papers previously identified. The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 46,463,117 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at start, all edges below the median normalized co-citation frequency were deleted. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster.  Using this approach, 22,232 clusters containing 84,591 nodes were generated, each containing less than 100 nodes. Clusters containing only 2 nodes were then discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned to the maximum edge weight of all edges between the nodes in these two clusters. Edges were arranged in descending order. The first pair of clusters was merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (Fig.~\ref{fig:merge_select}).

@@

As described in Methods, we performed 600 iterations to the 10,298 clusters further limited to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}). 
The 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

%%%%%%%%%
\section{Figures}
%%%%%%%%%

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.5]{graclus_comparison.pdf}
% figure caption is below the figure
\caption{Summary of DBLP data cross-matched with Scopus. 2,685,356 publications from DBLP were cross-matched with Scopus and then grouped by the 27 major subject areas in the ASJC (Scopus) classification. The largest number of publications are contributed by Computer Science; Mathematics; and Engineering and then by Social Sciences; Physics and Astronomy; and Biochemistry, Genetics, and Molecular Biology. Publications were further annotated with respect to being either articles \emph{(ar)} or conference proceedings \emph{(cp)}. For this dataset, the major subject areas of Computer Science; Engineering; and Mathematics contributed the most publications while Pharmacology, Toxicology, and Pharmaceutics; Nursing, Veterinary and Dentistry contributed the least.}
\label{fig:graclus_comparison}       
\end{figure}

@@

@@

\newpage
\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
 \includegraphics[scale=0.5]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{High-level disciplinary composition of publications in $\sim$2,000 clusters generated by variable-level clustering  of co-citation data.
%Variable Level Clustering. 
84,591 nodes from were matched to the 4 subject areas, 
27 major subject areas, and 334 minor subject areas in the Scopus ASJC classification. Since 82.5\% of these nodes were assigned more to than one minor subject category, 
we assigned fractional counts for the four subject areas that summed to 1 for each article. The figure shows a four way Venn Diagram~\cite{chen_venn_2011} illustrating the 
relative frequencies of publications from the Physical Sciences, Social Sciences, Life Sciences, and Health Sciences.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}

>>>>

\institute{Sitaram Devarakonda,  \at
              Netelabs, NET ESolutions Corporation, McLean, VA \\
              %Tel.: +123-45-678910\\
              %Fax: +123-45-678910\\
              \email{sitaram@nete.com}           %  \\
%             \emph{Present address:} of F. Author  %  if needed
           \and
              Dmitriy Korobskiy \at
              Netelabs, NET ESolutions Corporation, McLean, VA  \\
              \email{dk@nete.com}
           \and
           Tandy Warnow \at
              Dept of Computer Science, University of Illinois Urbana-Champaign, Champaign IL \\
              \email{warnow@illinois.edu}
           \and
           George Chacko \at
              Netelabs, NET ESolutions Corporation, McLean, VA \\
              \email{netelabs@nwete.com}
}

@@

Other classification systems are available such as the All Science Journal Classification (ASJC) developed and maintained by Scopus, the Web of Science research and categorical classifications from Clarivate Analytics, and the NSF classification system~\cite{nsf_classification}. All three rely on applying one or more journal-derived labels to articles.  Of these, Scopus and the Web of Science also include global bibliographies with citation links and proprietary unique identifiers for publications, some of which can be matched to digital object identifiers (DOIs). Applying labels designed to represent journals, to articles, can be problematic in terms of accuracy given broad scope in some journals,  and also because fields evolve faster than their journal classifications. [should we put in a comment about why article level labels provided by authors have limited use and that enforcement of controlled vocabularies varies by publisher?] Shu and colleagues recently noted in a comparative study of the Chinese Science Citation Database (CSCD) and the Web of Science, that 46\% of articles did not belong to the discipline of the journal they were published in~\cite{shu_comparing_2019}. Others have also discussed and critiqued disciplinary assignments using journal based classifications~\cite{wang_large-scale_2016,perianes-rodriguez_comparison_2017}. Article classification systems have been constructed, however, which escape some of the criticisms of journal-based classification~\cite{traag_louvain_2019,boyack_classification_2014,waltman_new_2012} but seem to be in the research domain rather than the domain of practical use [need to check on this].[May need another bridging sentence to the next paragraph].

@@

\subsection{Data}
 A stable release of the DBLP computer science bibliography~\cite{dblp_ref} consisting of 7,079,994 records was downloaded as dblp-2018-08-01.xml.gz. Publications were parsed from the XML source file and loaded into a PostgreSQL database. Just over 95\% of the publications within were published after 1996. We have previously parsed the Scopus dataset of over 72,000,000 publications into a custom schema as part of implementing a larger data platform for research evaluation~\cite{GithubERNIE2019}. Records in the DBLP dataset were matched to Scopus identifiers using digital object identifiers (DOIs). This procedure resulted in a dataset of 2,685,356 DBLP publications with Scopus identifiers where 1,278,322 (47.6\%) were labeled as article and 1,407,034 (52.4\%) as conference proceedings. References cited by these publications were then extracted from Scopus (7,129,006 records), resulting in a total of 8,000,411 nodes (publications and references) and 44,296,381 edges that represent citations within the dataset. This dataset is referred to as \emph{comp} (see Fig~\ref{fig:ar_cp_annotation}).

@@

In our implementation of variable level clustering, we first calculated the number of citations accumulated across Scopus for all 2,685,356 articles in \emph{comp}. After discarding those publications without any citations, we restricted further analysis to highly cited articles, those in the 90th percentile of citations or higher, resulting in a dataset of size 212,311. We then identified 4,318,305 publications in Scopus that cite these 212,311 articles from \emph{comp}. For each of the 4,318,305 citing publications in turn, all possible ${n \choose 2}$ reference pairs were generated, where $n$ is the number of references in a publication. The cited reference pairs were restricted to those in the set of 212,311 highly cited papers previously identified. The frequency of these co-cited pairs was then computed across the dataset  and normalized using Salton's cosine formula~\cite{salton_citation_1979}. A total of 46,463,117 unique co-cited pairs were thus obtained. These data were represented in a graph where each node was a publication and the weighted edge between the pair was the normalized co-citation frequency. \par

We applied variable clustering with initial parameters of threshold ($t$) = median normalized co-citation frequency (quantile=0.5), increment $i = 0.1$, and maximum cluster size, $mcs=200$. Thus, at start, all edges below the median normalized co-citation frequency were deleted. Clusters were formed by assembling connected components from each co-cited pair beginning with the heaviest edge weight. Clusters below size 100 were retained and any cluster larger than 200 nodes was carried over to the next round. The threshold, $t$, was then incremented by 0.1 and the process repeated while progressively incrementing $t$.  We used a bi-phasic approach where in which $t$ ranged from 0.5-0.9, after which $i$ was reduced to 0.01 for the range $0.9 \leq t \leq 0.99$. A final threshold of $t$=0.999 was applied to break the single remaining large cluster.  Using this approach, 22,232 clusters containing 84,591 nodes were generated, each containing less than 100 nodes. Clusters containing only 2 nodes were discarded bringing the total number of clusters down to 10,298.

Agglomerative clustering was then performed on these 10,298 clusters to generate higher-order clusters. Briefly, each cluster was now treated as a node and the edge weight between two clusters was assigned to the maximum edge weight of all edges between the nodes in the two precursor clusters. Edges were arranged in descending order. The first pair of clusters was merged and its edge weight with other interacting clusters was recalculated. All edges were then re-ordered as before and the next pair of clusters was merged. The process was halted after 600 rounds to prevent single large clusters being generated (Fig.~\ref{fig:merge_select}).

@@

As described in Methods, we performed 600 iterations of agglomerative clustering to the 10,298 clusters. To focus on major clusters, we restricted agglomeration to clusters containing at least 10 nodes, which resulted in 666 clusters. Further agglomeration resulted in the development of a single large cluster (Fig~\ref{fig:merge_select}).  Consequently, the 20 largest clusters were used for further analysis and comparison with clusters generated by direct citation. The largest of these 20 clusters contained 1,276 nodes and the smallest contained 136.

@@

%%%%%%%%%
\section*{Figures}
%%%%%%%%%

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
  \includegraphics[scale=0.4]{graclus_comparison.pdf}
% figure caption is below the figure
\caption{Conductance Measurements of Clusters Generated by Graclus.  2,685,356 DBLP publications, 7,129,006 cited references, and 44,296,381 citations were clustered using Graclus into 18 (grac\_18), 20 (grac\_20), or 22 (grac-22) clusters. Conductance, $\phi(S)$, was measured for these clusters considering only the edges between publications using the formula: $\phi(S)=|\partial(S)|/min(vol(S),2m-vol(S)$, where $\partial(S)$ is the boundary (number of edges leaving a set), vol(S) is volume of a set of vertices as the sum of the degrees of the vertices in a set, and $m$ is the number of undirected edges in a set~\cite{shun_parallel_2016}.}
\label{fig:graclus_comparison_conductance}       
\end{figure}

@@

\begin{figure}[ht]
\centering
% Use the relevant command to insert your figure file.
% For example, with the graphicx package use
 \includegraphics[scale=0.5]{cocit_venn2.pdf}
% figure caption is below the figure
\caption{High-level disciplinary composition of publications in $\sim$2,000 clusters generated by variable-level clustering  of co-citation data.
%Variable Level Clustering. 
84,591 nodes from were matched to the 4 subject areas, 
27 major subject areas, and 334 minor subject areas in the Scopus ASJC classification. Since 82.5\% of these nodes were assigned more to than one minor subject category, 
we assigned fractional counts for the four subject areas that summed to 1 for each article. The figure shows a four way Venn Diagram~\cite{chen_venn_2011} illustrating the 
relative frequencies of publications from the Physical Sciences, Social Sciences, Life Sciences, and Health Sciences.}
\label{fig:cocitr_venn2}       % Give a unique label
\end{figure}

@@
